ML Estimation of Matrix t Distributions with Expectation Maximization

### 2017-01-05 paper.tex
Formulas editing.

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration is the set of real symmetric
positive definite matrices with dimension $p$, $A$ is a $p\times p$
nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
function, given in terms of the standard gamma function $\Gamma(\cdot)$
by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as
$|\cdot|$. We also define the multivariate digamma function
%
\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

@@

We will show that the matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
follows a matrix--$t$ distribution with degrees of freedom $n=\nu-p+1$. Here $S^{\frac{1}{2}}$ is the lower Cholesky factor of $S$, i.e., $S=S^{\frac{1}{2}}(S^{\frac{1}{2}})^t$. The Jacobian determinant of the transformation \eqref{eq:change} is
\begin{equation}
\left|\frac{\partial T}{\partial X}\right|=|S|^{-\frac{m}{2}}.\label{eq:jac}
\end{equation}

Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint PDF of $T$ and $S$ is given as
\begin{equation}
\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right).
\end{split}
\label{eq:jointpdf}
\end{equation}
Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
by integrating over $S$ in the above expression, using Eq.~\eqref{eq:intmultgammadef} and slightly
rearranging the terms. In particular noting that
$\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times
p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$.

@@

The expectation maximization formalism entails calculating 
\begin{equation}
E=\sum_{i}^{N}\frac{\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})dS}{\textrm{tpdf}(T,M,\Sigma,\Omega,n)}\label{eq:expdef}
\end{equation}
where $\{T_{1}...T_{N}\}$ is an input set of data from which we wish
to estimate the maximum likelihood parameters of the distribution.
We omit writing explicitly the dependence of $E$ on all the parameters,
i.e. $E\equiv E(T_{i},S,M,\Sigma,\Omega,n,M_{t},\Sigma_{t},\Omega_{t},n_{t})$.
The values of the new guesses are found by computing the maximum with
respect to the parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$.

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\begin{equation}
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\end{equation}
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In order to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve Eq.~\ref{eq:firstlogpart} by substituting this result directly.

In order to compute the second nontrivial log term,
\begin{equation}
\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))dS
\end{equation}
we note that, due to the linearity of the trace, the result is equivalent
to solving a matrix valued integral
\begin{equation}
\textrm{tr}\left[\left(\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)SdS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
\end{equation}
The enclosed integral also follows from Eq.~\ref{eq:intmultgammadef},
after taking the matrix derivative with respect to $A$. Using, for
symmetric $S$
\begin{equation}
\frac{d}{dA}\det A=(\det A)A^{-1}\label{eq:detderivative}
\end{equation}
and
\begin{equation}
\frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
\end{equation}
we have 
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}SdS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
\end{equation}
from where again, we can obtain the result by direct substitution.

Combining results of the two integrals together, and with adequate
rearrangements we obtain the integrated form of Eq.~\ref{eq:expdef}:
\begin{equation}
\begin{split}E=\sum_{i}^{N}n_{t}\psi_{p}\left(\frac{1}{2}(m+n+p-1)\right)-2\log\Gamma_{p}\left(\frac{1}{2}(n_{t}+p-1)\right)+p\log\det\Omega_{t}^{-1}\\
-n_{t}\log\det\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)+(n_{t}+p-1)\log\det\Sigma_{t}\\
-(m+n+p-1)\textrm{tr}\left(\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left((T_{i}-M_{t})\Omega_{t}^{-1}(T_{i}-M_{t})^{\textrm{t}}+\Sigma_{t}\right)\right)
\end{split}
\label{eq:fsimp}
\end{equation}
The maximization formulas are obtained setting the derivatives of
this equation to zero.

@@

For $\Omega_{t}$ we obtain:
\begin{equation}
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\end{equation}
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\begin{equation}
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\begin{equation}
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}

>>>>

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}\,dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration is the set of real symmetric
positive definite matrices with dimension $p$, $A$ is a $p\times p$
nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
function, given in terms of the standard gamma function $\Gamma(\cdot)$
by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as
$|\cdot|$. We also define the multivariate digamma function
%
\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

@@

We will show that the matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
follows a matrix--$t$ distribution with degrees of freedom $n=\nu-p+1$. Here $S^{\frac{1}{2}}$ is the lower Cholesky factor of $S$, i.e., $S=S^{\frac{1}{2}}(S^{\frac{1}{2}})^t$. The Jacobian determinant of the transformation \eqref{eq:change} is
\begin{equation}
\left|\frac{\partial T}{\partial X}\right|=|S|^{-\frac{m}{2}}.\label{eq:jac}
\end{equation}
Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint density of $T$ and $S$ is given as
\begin{equation}
\begin{split}j(T,S;\left\{M,\Sigma,\Omega,n\right\})=\frac{\pi^{-\frac{mp}{2}}|\Sigma|^{\frac{n+p-1}{2}}|\Omega|^{\frac{n+m-2}{2}}|S|^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}\left[S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right]\right).
\end{split}
\label{eq:jointpdf}
\end{equation}

Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
by integrating over $S$ in the above expression, using Eq.~\eqref{eq:change} and slightly
rearranging the terms. In particular noting that
$|\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|=|\Sigma||I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|$.

@@

The expectation maximization formalism entails calculating 
\begin{equation}
E=\sum_{i=1}^{N}\frac{\int_{S\succ 0}j(T_{i},S,M,\Sigma,\Omega,n)\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\,dS}{f(T;\left\{M,\Sigma,\Omega,n\right\})},\label{eq:expdef}
\end{equation}
where $\{T_i\}_{i=1}^N$ is a dataset consists of matrix elements $T_i$ from which we wish to estimate the maximum likelihood parameters of the distribution. We omit writing explicitly the dependence of $E$ on all the parameters. The values of the new guesses are found by maximizing with respect to the parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$.

We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$ we have
\begin{equation}
\begin{split}\log j(T_{i},S;\left\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\right\})\sim n_{t}\log|S|+\textrm{tr}\left[S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\\
-n_{t}p\log2-p\log |\Omega_{t}|+(n_{t}+p-1)\log |\Sigma_{t}|-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1)),
\end{split}
\end{equation}
where the first two terms depend on $S$, and therefore we need to compute the corresponding integral, while for the rest, the integral is proportional to the PDF of the matrix--$t$ distribution, Eq.~\eqref{eq:matrixtpdf}.

To compute
\begin{equation}
n_{t}\int_{S\succ 0}j(T_{i},S;\left\{M,\Sigma,\Omega,n\right\})\log |S|\,dS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\eqref{eq:intmultgammadef} we have
\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}\left[SA\right]}|S|^{x}\log |S|\,dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}}(\log |A|+\gamma_{p}\left(\frac{1+p+2x}{2}\right)).\label{eq:intderivative}
\end{equation}
Equation~\eqref{eq:firstlogpart} can be computed by utilizing this result directly.

In order to compute the second nontrivial log term,
\begin{equation}
\int_{S\succ 0}j\left(T_{i},S,\left\{M,\Sigma,\Omega,n\right\}\right)\textrm{tr}\left[S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,dS,
\end{equation}
we note that due to the linearity of the trace the result is equivalent to solving a matrix valued integral
\begin{equation}
\textrm{tr}\left[\left(\int_{S\succ 0}j\left(T_{i},S,\left\{M,\Sigma,\Omega,n\right\}\right)S\,dS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
\end{equation}
The enclosed integral also follows from Eq.~\eqref{eq:intmultgammadef}, after taking the matrix derivative with respect to $A$. Using, for symmetric $S$
\begin{equation}
\frac{d}{dA}\det A=(\det A)A^{-1}\label{eq:detderivative}
\end{equation}
and
\begin{equation}
\frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
\end{equation}
we have 
\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}\left[SA\right]}|S|^{x}S\,dS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
\end{equation}
from where again, we can obtain the result by direct substitution.

Combining results of the two integrals together with adequate rearrangements we obtain the integrated form of Eq.~\eqref{eq:expdef}
\begin{equation}
\begin{split}E=\sum_{i}^{N}n_{t}\psi_{p}\left(\frac{1}{2}(m+n+p-1)\right)-2\log\Gamma_{p}\left(\frac{1}{2}(n_{t}+p-1)\right)+p\log|\Omega_{t}|^{-1}\\
-n_{t}\log|(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma|+(n_{t}+p-1)\log |\Sigma_{t}|\\
-(m+n+p-1)\textrm{tr}\left(\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left((T_{i}-M_{t})\Omega_{t}^{-1}(T_{i}-M_{t})^{\textrm{t}}+\Sigma_{t}\right)\right)
\end{split}
\label{eq:fsimp}
\end{equation}
The maximization formulas are obtained setting the derivatives of this equation to zero.

@@

For $\Omega_{t}$ we obtain:
\begin{equation}
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\end{equation}
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\begin{equation}
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\det\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\begin{equation}
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
